Solving \\ the \\ 15-Puzzle \\ \& \\ Rubik's Cube

### 2022-09-08 latex/Appendices/Examples.tex
update

\chapter{Examples} % Main chapter title

@@

%-----------------------------------
%	SECTION 1
%-----------------------------------
\section{Puzzles}
Let me start by showing how to construct puzzles, using the Puzzle factory. Notice that in order to run a learner or solver of any kind (assuming of course that they are meant to work on the puzzle type in question), we can just use the exact same code, simply specifying \textit{puzzle\_type} and the expected parameters to construct a puzzle. The factories will just pick up the relevant parameters indicating the puzzle type and dimension and everything should work seamlessly.
\\
\\
For instance, let us create a 5x6 \textbf{SP}, shuffle it a thousand times, and print it:
\afblue
\paragraph{}{\textbf{python code -- sliding puzzle construction}}
\begin{python}
####################################################################
from rubiks.puzzle.puzzle import Puzzle
####################################################################
puzzle_type=Puzzle.sliding_puzzle
n=5
m=6
nb_moves=1000
print(Puzzle.factory(**globals()).apply_random_moves(nb_moves))
####################################################################
\end{python}
\black
The output from the above code snippet will look like (subject to randomness):

@@

%-----------------------------------
%	SECTION 2
%-----------------------------------
\section{Learners}

%-----------------------------------
%	SUBSECTION 2.1
%-----------------------------------
\subsection{Perfect Learner}
\label{PLSS}

@@

%-----------------------------------
%	SUBSECTION 2.2
%-----------------------------------
\subsection{Deep Learner}
\label{DLSS}
The DeepLearner, also discussed in details in \ref{sec:codelearners}, needs some training data. It could in principle of course be trained on any data, not necessarily optimal data (i.e. generated by an optimal solver). Here however, I make use of the TrainingData class to generate 100 sequences of fully solved 5x2 \textbf{SP} via A{*} with Manhattan++.

@@

%-----------------------------------
%	SUBSECTION 2.3
%-----------------------------------
\subsection{Deep Reinforcement Learner}
\label{DRLSS}

@@

\section{Solvers}

%-----------------------------------
%	SUBSECTION 3.1
%-----------------------------------
\subsection{Blind search}
First let me start with some of the blind search algorithms, which I only have been able to use on small dimension SP. They quickly become too memory hungry to be practical on anything but the smallest puzzles.
%-----------------------------------
%	SUBSECTION 3.1.1
%-----------------------------------
\subsubsection{BFS}
\label{BFSSS}
The following example shows how to use Breadth First Search to solve a (n=3, m=3) SP which we do not \textit{shuffle} too much.

@@

%-----------------------------------
%	SUBSECTION 3.1.2
%-----------------------------------
\subsubsection{DFS}
\label{DFSSS}

@@

\subsection{Naive Sliding Puzzle Solver}

@@

%-----------------------------------
%	SUBSECTION 3.3
%-----------------------------------
\subsection{Kociemba}

@@

%-----------------------------------
%	SUBSECTION 3.4
%-----------------------------------
\subsection{A*}
\label{ASSS}

\subsubsection{Manhattan heuristic}

@@

\subsubsection{Perfect Heuristic}
To run A{*} with a perfect heuristic, we just need to specify the heuristic type as such, and set the parameter \textit{model\_file\_name} to point to a pre-recorded database populated by the PerfectLearner (see earlier section \ref{PLSS}).

@@

\subsubsection{Deep Learning Heuristic}
\subsubsection{Deep Reinforcement Learning Heuristic}
\subsubsection{Deep Q Learning Heuristic}

>>>>

\Appendix{Examples} % Main chapter title

@@

%-----------------------------------
%	SECTION 1
%-----------------------------------
\Section{Puzzles}
Let me start by showing how to construct puzzles, using the Puzzle factory. Notice that in order to run a learner or solver of any kind (assuming of course that they are meant to work on the puzzle type in question), we can just use the exact same code, simply specifying \textit{puzzle\_type} and the expected parameters to construct a puzzle. The factories will just pick up the relevant parameters indicating the puzzle type and dimension and everything should work seamlessly.
\\
\\
For instance, let us create a 5x6 \textbf{SP}, shuffle it a thousand times, and print it:
\afblue
\paragraph{}{\textbf{python code -- sliding puzzle construction}}
\begin{python}
####################################################################
from rubiks.puzzle.puzzle import Puzzle
####################################################################
puzzle_type=Puzzle.sliding_puzzle
n=5
m=6
nb_moves=1000
print(Puzzle.factory(**globals()).apply_random_moves(nb_moves))
####################################################################
\end{python}
\black
The output from the above code snippet will look like (subject to randomness):

@@

%-----------------------------------
%	SECTION 2
%-----------------------------------
\Section{Learners}

%-----------------------------------
%	SUBSECTION 2.1
%-----------------------------------
\Subsection{Perfect Learner}
\label{PLSS}

@@

%-----------------------------------
%	SUBSECTION 2.2
%-----------------------------------
\Subsection{Deep Learner}
\label{DLSS}
The DeepLearner, also discussed in details in \ref{sec:codelearners}, needs some training data. It could in principle of course be trained on any data, not necessarily optimal data (i.e. generated by an optimal solver). Here however, I make use of the TrainingData class to generate 100 sequences of fully solved 5x2 \textbf{SP} via A{*} with Manhattan++.

@@

%-----------------------------------
%	SUBSECTION 2.3
%-----------------------------------
\Subsection{Deep Reinforcement Learner}
\label{DRLSS}

@@

\Section{Solvers}

%-----------------------------------
%	SUBSECTION 3.1
%-----------------------------------
\Subsection{Blind search}
First let me start with some of the blind search algorithms, which I only have been able to use on small dimension SP. They quickly become too memory hungry to be practical on anything but the smallest puzzles.
%-----------------------------------
%	SUBSECTION 3.1.1
%-----------------------------------
\Subsubsection{BFS}
\label{BFSSS}
The following example shows how to use Breadth First Search to solve a (n=3, m=3) SP which we do not \textit{shuffle} too much.

@@

%-----------------------------------
%	SUBSECTION 3.1.2
%-----------------------------------
\Subsubsection{DFS}
\label{DFSSS}

@@

\Subsection{Naive Sliding Puzzle Solver}

@@

%-----------------------------------
%	SUBSECTION 3.3
%-----------------------------------
\Subsection{Kociemba}

@@

%-----------------------------------
%	SUBSECTION 3.4
%-----------------------------------
\Subsection{A*}
\label{ASSS}

\Subsubsection{Manhattan heuristic}

@@

\Subsubsection{Perfect Heuristic}
To run A{*} with a perfect heuristic, we just need to specify the heuristic type as such, and set the parameter \textit{model\_file\_name} to point to a pre-recorded database populated by the PerfectLearner (see earlier section \ref{PLSS}).

@@

\Subsubsection{Deep Learning Heuristic}
\Subsubsection{Deep Reinforcement Learning Heuristic}
\Subsubsection{Deep Q Learning Heuristic}

### 2022-09-08 latex/Chapters/Conclusion.tex
update

TBW

>>>>

\Section{Learning}
Learnt about puzzles, and also even took the time to learn a couple of sequences of moves that are sufficient to solve \textbf{RC}, 2x2x2 in labout one minute, 3x3x3 in about 2 minutes.

\Section{Implementation}
Implemented a bunch of solvers. Ideally should do in C++ for speed and multithreaded (esp for MCTS) but time was limited so I prioritized breadth over depth.

\Section{Deep Reinforcement/Q Learning}
Quite impressive, without too much effort, how well they perform. However, I feel to really get more out of them (higher dimensions in particular) would need careful tuning of the convergence criteria.

\Section{Professional issues}
See \cite{ManyWorlds}
See \cite{https://doi.org/10.1111/1758-5899.12718} and \cite{Bostrom2014}

### 2022-09-08 latex/Chapters/Puzzles.tex
update

\section{Sliding Puzzle}

%-----------------------------------
%	SUBSECTION 1.1
%-----------------------------------
\subsection{History - the 15-Puzzle}

@@

\begin{figure}[H]
\centering
\includegraphics[scale=0.3]{./Figures/coiled15puzzle}
\includegraphics[scale=0.3]{./Figures/coiled15puzzle2.png}
\decoRule
\caption[Coiled 15-puzzle]{Coiled 15-puzzle}
\end{figure}

@@

%-----------------------------------
%	SUBSECTION 1.2
%-----------------------------------
\subsection{Search Space \& Solvability}

@@

%-----------------------------------
%	SUBSECTION 1.3
%-----------------------------------
\subsection{Optimal Cost \& God's Number}

@@

%-----------------------------------
%	SECTION 2
%-----------------------------------
\section{Rubik's Cube}

\subsection{History -- from Erno Rubik to speed cubing competitions}
The second puzzle I will focus on is the well known Rubik's Cube (\textbf{RC}), invented in 1974 by Erno Rubik, originally known under the name of \textit{Magic Cube} (see \cite{RubiksWiki}). Since it was commercialized in 1980, it has kown a worldwide success: countless people are playing with the famous original 3x3x3 cube, as well as with many variants of it, more or less difficult and contrived. Competitions nowadays bring together \textit{speedcubers} who have trained for years and memorized various algorithms to solve the Rubik's in astonishly effective and fast times (literally in seconds for the best ones). Interestingly, some of the principles used by speedcubers generalise to different dimensions. As an example, it is quite impressive to watch \textit{Cubastic} solve a 15x15x15 cube in (only) two and a half hours! (see \cite{151515Rubiks}). The key insight to solving high dimensional \textbf{RC}s (short obviously of coming up with specific and sophisticated strategies) is the following: if we manage to group all the inside (i.e. not located on a corner or an edge of the cube) cubies by colour, and then all the non-corner edges on a given tranche of a face by colour, we are essentially reducing the cube to a 3x3x3, since no face rotation will ever undo the inside cubies and non-corner-edge cubies' grouping. For instance, the following picture shows faces of my own 4x4x4 and 3x3x3 cubes which I have arranged to be in identical configurations.
\\
\\
It is quite interesting to note that the difficuly of hand-solving cubes does not grow nearly as much with increasing dimensionality (most of the time difference comes rather from the difficulty of handling and manipulating the cubes of high dimensions at speed).

\begin{figure}[H]
\centering
\includegraphics[scale=0.3]{./Figures/RC333444}
\decoRule
\caption[Reducing RC]{Reducing a 4x4x4 \textbf{RC} to a 3x3x3 -- home made picture}
\end{figure}

@@

\subsection{Conventions: faces, moves and configurations}
In my code as well as in this write-up, I follow the conventions and notations of the \textbf{RC} literature and community as much as possible, I will refer to the faces as F (front), B (back), U (up), D (down), L (left) and R (right). The cubes' tiles are of 6 different colors, and without loss of generality (though the interested reader is referred to appendix \ref{sec:Kociemba} for some subtlety around that), we can consider that these colors are also called F, B, U, D, L and R, and consider the goal state as one where, up to rotations of the whole cube, the color's name match the faces (color F on face F, etc). The Rubik's has, as one would expect, an extremely large state space. I will compute the state space size for the 3x3x3 and 2x2x2 cubes in the next two subsections and discuss conditions under which  a cube is solvable.
\\
It is common in \textbf{RC} jargon to refer to moves  by the name of the face that is rotated clock-wise (when facing it), and by adding a prime \textit{'} for counter-clock-wise rotations. Sometimes people use a number after the face to indicate repeated similar moves, though I will count in my code that as 2 moves, since obviously e.g. U2 = U U = U' U'. As another example: F B' F2 means rotating the front face clock-wise, the back face counter-clock-wise, followed by the front face twice. Some people refer to the convention I have chosen as the \textit{quarter turn metric}, and obvioulsy the length of solutions is affected by which convention/metric we use.
\\
Finally, let me make a remark on what I consider in this project to be distinct configurations of the Rubik's cube. Since I am interested in experimenting with, among other methods, \textbf{DRL} and \textbf{DQL}, and that they are usually presented as methods which learn \textit{without-human-knowledge}, I thought it would be a bit akin to cheating to consider that two cubes equivalent by full space rotation are one and the same and represent them as such in my code. Ideally, we would rather let the algorithms learn (if they are able to) about symmetries. For instance, the two following cubes (which are clearly just a rotation away or equivalently a change-in-observer away from one another) are the same for us humans. I however represent them as different tensors in my code, and this is up to the solvers to make sense of things.

@@

\subsubsection{3x3x3 Rubik's -- search space, solvability \& God's number}

@@

\subsubsection{2x2x2 Rubik's -- search space, solvability}

@@

%-----------------------------------
%	SECTION 3
%-----------------------------------
\section{Solving puzzles without human knowledge}
\label{sec:WHK}

>>>>

\Section{Sliding Puzzle}

%-----------------------------------
%	SUBSECTION 1.1
%-----------------------------------
\Subsection{History - the 15-Puzzle}

@@

\begin{figure}[H]
\centering
\includegraphics[scale=0.3]{./Figures/coiled15puzzle}
\includegraphics[scale=0.3]{./Figures/coiled15puzzle2.png}
\caption[Coiled 15-puzzle]{Coiled 15-puzzle}
\end{figure}

@@

%-----------------------------------
%	SUBSECTION 1.2
%-----------------------------------
\Subsection{Search Space \& Solvability}

@@

%-----------------------------------
%	SUBSECTION 1.3
%-----------------------------------
\Subsection{Optimal Cost \& God's Number}

@@

%-----------------------------------
%	SECTION 2
%-----------------------------------
\Section{Rubik's Cube}

\Subsection{History -- from Erno Rubik to speed cubing competitions}
The second puzzle I will focus on is the well known Rubik's Cube (\textbf{RC}), invented in 1974 by Erno Rubik, originally known under the name of \textit{Magic Cube} (see \cite{RubiksWiki}). Since it was commercialized in 1980, it has kown a worldwide success: countless people are playing with the famous original 3x3x3 cube, as well as with many variants of it, more or less difficult and contrived. Competitions nowadays bring together \textit{speedcubers} who have trained for years and memorized various algorithms to solve the Rubik's in astonishly effective and fast times (literally in seconds for the best ones). Interestingly, some of the principles used by speedcubers generalise to different dimensions. As an example, it is quite impressive to watch \textit{Cubastic} solve a 15x15x15 cube in (only) two and a half hours! (see \cite{151515Rubiks}). The key insight to solving high dimensional \textbf{RC}s (short obviously of coming up with specific and sophisticated strategies) is the following: if we manage to group all the inside (i.e. not located on a corner or an edge of the cube) cubies by colour, and then all the non-corner edges on a given tranche of a face by colour, we are essentially reducing the cube to a 3x3x3, since no face rotation will ever undo the inside cubies and non-corner-edge cubies' grouping. For instance, the following picture shows faces of my own 4x4x4 and 3x3x3 cubes which I have arranged to be in identical configurations.
\\
\\
It is quite interesting to note that the difficuly of hand-solving cubes does not grow nearly as much with increasing dimensionality (most of the time difference comes rather from the difficulty of handling and manipulating the cubes of high dimensions at speed).

\begin{figure}[H]
\centering
\includegraphics[scale=0.3]{./Figures/RC333444}
\caption[Reducing RC]{Reducing a 4x4x4 \textbf{RC} to a 3x3x3 -- home made picture}
\end{figure}

@@

\Subsection{Conventions: faces, moves and configurations}
In my code as well as in this write-up, I follow the conventions and notations of the \textbf{RC} literature and community as much as possible, I will refer to the faces as F (front), B (back), U (up), D (down), L (left) and R (right). The cubes' tiles are of 6 different colors, and without loss of generality (though the interested reader is referred to appendix \ref{sec:Kociemba} for some subtlety around that), we can consider that these colors are also called F, B, U, D, L and R, and consider the goal state as one where, up to rotations of the whole cube, the color's name match the faces (color F on face F, etc). The Rubik's has, as one would expect, an extremely large state space. I will compute the state space size for the 3x3x3 and 2x2x2 cubes in the next two subsections and discuss conditions under which  a cube is solvable.
\\
It is common in \textbf{RC} jargon to refer to moves  by the name of the face that is rotated clock-wise (when facing it), and by adding a prime \textit{'} for counter-clock-wise rotations. Sometimes people use a number after the face to indicate repeated similar moves, though I will count in my code that as 2 moves, since obviously e.g. U2 = U U = U' U'. As another example: F B' F2 means rotating the front face clock-wise, the back face counter-clock-wise, followed by the front face twice. Some people refer to the convention I have chosen as the \textit{quarter turn metric}, and obvioulsy the length of solutions is affected by which convention/metric we use.
\\
Finally, let me make a remark on what I consider in this project to be distinct configurations of the Rubik's cube. Since I am interested in experimenting with, among other methods, \textbf{DRL} and \textbf{DQL}, and that they are usually presented as methods which learn \textit{without-human-knowledge}, I thought it would be a bit akin to cheating to consider that two cubes equivalent by full space rotation are one and the same and represent them as such in my code. Ideally, we would rather let the algorithms learn (if they are able to) about symmetries. For instance, the two following cubes (which are clearly just a rotation away or equivalently a change-in-observer away from one another) are the same for us humans. I however represent them as different tensors in my code, and this is up to the solvers to make sense of things.

@@

\Subsection{3x3x3 Rubik's -- search space, solvability \& God's number}

@@

\Subsection{2x2x2 Rubik's -- search space, solvability}

@@

%-----------------------------------
%	SECTION 3
%-----------------------------------
\Section{Solving puzzles without human knowledge}
\label{sec:WHK}
